Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-26 manu-draft.Rmd
revisions from CA comments

From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L. Cyanobacteria blooms (*Dapis* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.  Blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.  Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years.

@@

Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.  The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].  Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13]. Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O). Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].  The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b]. There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks. Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).  Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].

Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.  In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds. At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.  Unlike past discharges from the site, HRK was authorized to release untreated wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.  This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.  From March 30th to April 9th, approximately 814 million liters (215 million gallons) of wastewater were released to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point. This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.

@@

## Timeline of events from April 2021

A general narrative of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgam)b).  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

From April through September 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).  Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others). The percentage of observations outside of the normal range varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).  Physical parameters (salinity, temperature) were generally within range over the study period. Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the discharge measurements in Table \@ref(tab:stktab). Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

@@

Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.  *Dapis sp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change dramatically from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

@@

## Nutrient cycling

>>>>

Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.  From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).  An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.  An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L. Cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.  Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.  By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release of wastewater from Piney Point are abnormal relative to historic conditions in Tampa Bay.

@@

Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].  The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O). Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].  The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b]. Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain wastewater on site. There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks. Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].

Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.  Ecosystem management paradigms for estuaries are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.  The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14]. Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].  Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].  Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17]. The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.  The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement. The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).  Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.  Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.  Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].

Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.  In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility. At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.  Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay, particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L). From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.  Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].

This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.  The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region. A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.  We focus primarily on the perspective of the Tampa Bay Estuary Program (TBEP) in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions. The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.

@@

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. These areas (Areas 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals). Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point. This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.

@@

## Timeline of 2021 biological response events

A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline). After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.  Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April chlorophyll concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L. The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.  On April 20th, *K. brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.  Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.  *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.  By June 27th, fish kill reports attributed to red tide increased with *K. brevis* cell concentrations in lower and middle Tampa Bay.  The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.  This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St. Petersburg and Tampa. Concentrations of *K. brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.  Conditions were relatively stable in August and September compared to the months prior. A quantitative description of these events follows.

@@

Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).  Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay. Similar differences for total phosphorus, total nitrogen, and chlorophyll-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.  These contrasts provided a context for the monitoring data collected in 2021.

From April through September 2021, `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).  Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others). The percentage of observations outside of the normal range varied by location and parameter.  For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively. Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.  Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.  Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).  Physical parameters (salinity, temperature) were generally within range over the study period. Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the discharge measurements in Table \@ref(tab:stktab). Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).

@@

Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex). Transect S3T6 is located less than one kilometer to the north of Port Manatee. Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.  Red macroalgae were present in high abundances from April to May.  *Dapis sp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.  *Dapis sp.* persisted through June and July, but not was observed after July 20th.  Green macroalgae taxa were first observed in July, although at generally low abundances. Red macroalgae were the dominant taxa by the end of September. Overall abundance of seagrass did not change dramatically from April 22nd through September. The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.

@@

## Potential nutrient cycling
